\documentclass[1p]{elsarticle}
\usepackage{amssymb,latexsym}
\usepackage{enumerate}


%%%% Put my macros here:
\input{myMacros.sty}

%% A numbered theorem with a fancy name:

\newtheorem{mainthm}[thm]{Main Theorem}

%% Numbered objects of "non-theorem" style (text roman):

\theoremstyle{definition}
\newtheorem{defin}[thm]{Definition}
\newtheorem{rem}[thm]{Remark}
\newtheorem{exa}[thm]{Example}

\begin{document}


%%%%%%%%%%%


\title{A quantitative nonlinear strong ergodic theorem for Hilbert spaces}
\author{Pavol Safarik}
\ead{pavol.safarik@gmail.com}
\address{Department of Mathematics, Technische Universit\"at Darmstadt, Schlossgartenstra{\ss}e 7, 64289 Darmstadt, Germany}

\date{}

\begin{abstract}
We give a quantitative version of a strong nonlinear ergodic theorem for (a class of possibly even discontinuous) 
selfmappings of an arbitrary subset of a Hilbert space due to R.~Wittmann and 
outline how the existence of uniform bounds in such quantitative formulations of ergodic theorems
can be proved by means of a general logical metatheorem.
In particular these bounds depend neither on the operator nor on the initial point.
Furthermore, we extract such
uniform bounds in our quantitative formulation of Wittmann's theorem, implicitly using the 
proof-theoretic techniques on which the metatheorem is based. However, we present our result
and its proof in analytic terms without any reference to logic as such. Our bounds turn out to involve
nested iterations of relatively low computational complexity. While in theory these kind
of iterations ought to be expected, so far this seems to be the first occurrence
of such a nested use observed in practice. %unprecedented form.
\end{abstract}

\begin{keyword}
Proof mining \sep uniform bounds \sep functionals of finite type \sep nonlinear ergodic theory \sep strong convergence \sep Ces{\`a}ro means \sep hard analysis.\\
\MSC[2010]{03F10, 47H25, 37A30}.
\end{keyword}


\maketitle

\input{intro.tex}

\input{aLogicalMetatheorem.tex}


\input{WittmannReallyArithmetical.tex}

\input{terms.tex}

\input{witnessProof.tex}

\input{Lemmas.tex}

\subsection*{Acknowledgements}
I wish to thank Ulrich Kohlenbach for helpful discussions and for suggesting the topic.
This research was supported by the German Science Foundation (DFG Project KO 1737/5-1). 

\input{biblio.tex}

\end{document}
